---
title: "HD Code"
Author: "Muhammad Elsadany"
---


```{r}
# load libraries
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(DESeq2)
library(pheatmap)
library(boxplotdbl)
library(gplots)
library(org.Hs.eg.db)
```



#Data retrieval part
```{r}
HDData <- read.csv("data/HD_Counts.csv", header=T, check.names=FALSE, row.names=1)
HDmeta <- read.csv("data/HD_Meta.csv", header=T, check.names = FALSE,  row.names=1)
```


#Check step
```{r}

### Check that sample names match in both files
all(colnames(HDData) %in% rownames(HDmeta))
all(colnames(HDData) == rownames(HDmeta))

```



#DESeq Object for comparing HD VS. Control
```{r}
## Create DESeq2Dataset object
HDdds <- DESeqDataSetFromMatrix(countData = HDData, colData = HDmeta, design = ~ SampleType)
#View(counts(HDdds))
## Run analysis
HDdds <- DESeq(HDdds)
```



```{r}
print(HDdds)

```



#generating normalized counts
```{r}
HDdds2 <- estimateSizeFactors(HDdds)
HDnormalized_counts <- counts(HDdds2, normalized=TRUE)
#write.csv(HDnormalized_counts, file="data/HDnormalized_counts.csv")

```


#contrast
```{r}
## Define contrasts, extract results table, and shrink the log2 fold changes

#Define your disease group first, then your control 

contrast_ <- c("SampleType", "HD", "Control")
resultsNames(HDdds)


res_table_unshrunken <- results(HDdds, contrast=contrast_, alpha = 0.05, lfcThreshold = 0)

res_table_shrunken <- lfcShrink(HDdds, contrast=contrast_, res=res_table_unshrunken, type="ashr")

```


# summary function

```{r}
## Summarize results
summary(res_table_unshrunken)

```

#saving results table
```{r}
res_table_unshrunken <- res_table_unshrunken %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
write.csv(res_table_unshrunken, file="results/HD_VS._control_res_table_unshrunken.csv")



res_table_shrunken <- res_table_shrunken %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
write.csv(res_table_shrunken, file="results/HD_VS._control_res_table_shrunken.csv")

```





#setting thresholds for DEGs
```{r}
### Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0

```


#subsetting table for DEGs values only
```{r}
sigOE <- res_table_unshrunken %>%
        filter(padj < padj.cutoff & abs(log2FoldChange) > lfc.cutoff)

#write.csv(sigOE, file="results/HD_VS._control_DEGslfc0.csv")
```




#converting ENSG of DEGs to GENE SYMBOL 
```{r}

df.ensemble.id <- sapply(sigOE$gene, function(x) strsplit(as.character(x),"\\.")[[1]][1])
  df.in <- cbind(df.ensemble.id, sigOE)

mapper <- mapIds(org.Hs.eg.db,keys = sigOE$gene, column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")

mapper.df <- as.data.frame(mapper)
mapper.df <- cbind(rownames(mapper.df), mapper.df)
names(mapper.df) <- c("df.ensemble.id", "symbol")
df.out <- merge(df.in, mapper.df ,by="df.ensemble.id", all.x=TRUE)


df.out <- df.out[-1]
df.out <- df.out[ ! is.na(df.out$symbol),]



```

#saving DEGs after conversion
```{r}
write.csv(df.out, file="results/HD_VS._control_ConvertedDEGslfc0.csv")

```











#creating tibble objects 
```{r}
# Create tibbles including row names
HD_meta <- HDmeta %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()
        
HDnormalized_counts <- HDnormalized_counts %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()

```






#mutating res_table with threshold
```{r}
## Obtain logical vector where TRUE values denote padj values < 0.05 and fold change > 0 in either direction

res_table_shrunken_2 <- res_table_shrunken %>% 
                  mutate(threshold_OE = abs(log2FoldChange) > 0)

```

#volcano plot
```{r}
## Volcano plot
jpeg(file="results/HD_VS._control_Volcano_plot.png")
ggplot(res_table_shrunken_2) +
        geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold_OE)) +
        ggtitle("HD VS. Control") +
        xlab("log2 fold change") + 
        ylab("-log10 adjusted p-value") +
        #scale_y_continuous(limits = c(0,50)) +
        theme(legend.position = "none",
              plot.title = element_text(size = rel(1.5), hjust = 0.5),
              axis.title = element_text(size = rel(1.25)))  


```







#heatmap for DEGs
```{r}
### Annotate our heatmap (optional)
annotation2 <- HD_meta %>% 
	dplyr::select(samplename, SampleType) %>% 
	data.frame(row.names = "samplename")

vsd <- vst(HDdds)

### Set a color palette
heat_colors2 <- brewer.pal(4, "YlOrRd")


genes <- sigOE$gene

### Run pheatmap

jpeg(file="results/HD_VS._control_DEGsHeatMaplfc0_MS.png")
pheatmap(assay(vsd)[genes, ],
         color = heat_colors2, 
         cluster_rows = T, 
         cluster_cols = F,
         show_rownames = F,
         show_colnames = F,
         annotation = annotation2, 
         border_color = NA, 
         fontsize = 10, 
         scale = "row", 
         fontsize_row = 6, 
         height = 400)
#dev.off()

```
















